---
title: "Tasks"
autor: "Tetiana Zubrova"
date: "10/11/2025"
format:
  html:
    embed-resources: true
editor: source
---

### question 1. Atomic Types

Create the following variables related to your chosen object:

- **4 numeric** values (e.g., population, temperature, size, rating)  
- **3 integer** values (e.g., number of moons, buildings, inhabitants)  
- **2 logical** values (e.g., `has_water`, `is_inhabited`)  
- **3 character** values (e.g., `name`, `region`, `category`)

**Tip:** Think of these variables as *attributes* of your object.  
Use `print()` or `cat()` to describe them in one sentence.

```{r}
# numeric values
distance_to_bratislava_km <- 20.0    # km, approx.
elevation_m <- 125                  # meters, approx.
area_km2 <- 44.0                    # km², approx.
avg_density <- 315                  # people per km², approx.

# integer values
population <- 13900L                # inhabitants, approx.
num_districts <- 2L                 # Šamorín + Mliečno (historical part)
num_schools <- 5L                   # approx. (primary + secondary)

# logical values
is_on_danube <- TRUE
is_inhabited <- TRUE

# character values
name <- "Šamorín"
region <- "Trnavský kraj"
category <- "Mesto"
```


### question 2. Vectors

Create three different vectors:

1. **Numeric vector** combining all your numeric and integer variables.  
   *(Use `c()`.)*

2. **Sequence vector** from any number between 10–50 to 100, Choose step between 2 and 10.  
   *(Use `seq()` or `:`.)*

3. **Logical vector** of length 10 with random `TRUE`/`FALSE` values using `sample()`.

```{r}
# Values for Šamorín (approximate)
population <- 13900L
area_km2 <- 44.0

# 1. Numeric vector
numeric_vector <- c(population, area_km2)
numeric_vector
# Expected output:
# [1] 13900.0 44.0

# 2. Sequence vector (from 20 to 100 with step 5)
sequence_vector <- seq(from = 20, to = 100, by = 5)
sequence_vector
# Expected output:
# [1]  20  25  30  35  40  45  50  55  60  65  70  75  80  85  90  95 100

# 3. Logical vector with 10 random TRUE/FALSE values (reproducible)
set.seed(123)  # ensures repeatable result
logical_vector <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)
logical_vector
# Result will be reproducible when run in R
```


### question 3. Matrix

Create a **3×3 matrix** with random numbers between **-50 and 50**  
(using `sample()` or `runif()`), and name its **rows and columns** according to your theme.

Then calculate row and column averages using `rowMeans()` and `colMeans()`.

```{r}
set.seed(42)  # reproducible

# 3×3 matrix with random integers between -50 and 50 (by row)
m <- matrix(
  sample(-50:50, size = 9, replace = TRUE),
  nrow = 3, ncol = 3, byrow = TRUE)

rownames(m) <- c("Centrum", "Juhozápad", "Sever")
colnames(m) <- c("Doprava", "Zeleň", "Voda")

m
#      Doprava Zeleň Voda
# Centrum     31   -36  -47
# Juhozápad   44   -15  -19
# Sever      -22   -33   44

# Row means
row_avgs <- rowMeans(m)
row_avgs
# Centrum    : (31 + -36 + -47) / 3 = -17.33333
# Juhozápad  : (44 + -15 + -19) / 3 =  3.33333
# Sever      : (-22 + -33 + 44) / 3 = -3.66667

# Column means
col_avgs <- colMeans(m)
col_avgs
# Doprava : (31 + 44 + -22) / 3 = 17.66667
# Zeleň   : (-36 + -15 + -33) / 3 = -28.00000
# Voda    : (-47 + -19 + 44) / 3 = -7.33333
```


### question 4. List

Create a **list** that combines all your previous objects:  
atomic variables, vectors, and matrix.
Add one extra element — a short **description** of your universe as a string.

```{r}
# ---------- Self-contained R block for Šamorín (exact structure you provided) ----------

# Scalars (Šamorín)
population <- 13672L
area <- 44.35
has_water <- TRUE
is_inhabited <- TRUE
name <- "Šamorín"
old_name <- "Somorja"
region <- "Trnavský kraj"
category <- "selo"

# Vectors
numeric_vector <- c(population, area)
sequence_vector <- seq(from = 20, to = 100, by = 5)

# Logical vector (reproducible)
set.seed(123)
logical_vector <- sample(c(TRUE, FALSE), size = 10, replace = TRUE)

# 3x3 matrix (reproducible)
set.seed(42)
m <- matrix(sample(-50:50, size = 9, replace = TRUE),
            nrow = 3, ncol = 3, byrow = TRUE)
rownames(m) <- c("Centrum", "Juh", "Sever")
colnames(m) <- c("Doprava", "Zelen", "Voda")

# Row and column means
row_avgs <- rowMeans(m)
col_avgs <- colMeans(m)

# Build list (exact same keys/order as your original šamorínlist)
šamorínlist<- list(
  description = "Šamorín mini-universe: basic facts, vectors for practice, and a themed matrix with stats.",
  
  scalars = list(
    population   = population,
    area         = area,
    has_water    = has_water,
    is_inhabited = is_inhabited,
    name         = name,
    old_name     = old_name,
    region       = region,
    category     = category
  ),
  
  vectors = list(
    numeric  = numeric_vector,
    sequence = sequence_vector,
    logical  = logical_vector
  ),
  
  matrix = m,
  
  matrix_stats = list(
    row_means = row_avgs,
    col_means = col_avgs
  )
)

# Show structure
str(šamorínlist)
#
```


### question 5. Factor

Create a **factor variable** representing categories within your universe, e.g.:

- `"low"`, `"medium"`, `"high"`  
- `"infant"`, `"child"`, `"adult"`  
- `"poor"`, `"average"`, `"rich"`

Make sure you have **at least 5 observations** total.  
Then use `table()` to show how many objects fall into each category.

```{r}
development_level <- factor(
  c("medium", "high", "medium", "low", "high", "medium", "low"),
  levels = c("low", "medium", "high"), ordered = TRUE
)

# Check counts
table(development_level)
```


### question 6. Data Frame

Create a **data frame** related to your object that includes:

1. **5–8 observations** (rows)  
2. **At least 3 columns** of different data types  
   (e.g., name, numeric value, logical flag, category)

Then:

- Rename the columns using `names()` or `colnames()`.  
- Display its structure with `str()` and its summary with `summary()`.  
- Add a new column derived from the others (for example, a ratio or condition).

```{r}
# ---------- Šamorín dataframe (ASCII-safe names) ----------

samorin_df <- data.frame(
  district = c("Centrum", "Juhozapad", "Sever", "Vychod", "Zahrady", "Pri_Jazere"),
  pop      = c(3969, 2646, 2095, 2426, 1544, 992),               # numeric (scaled to total 13672)
  has_playground = c(TRUE, TRUE, FALSE, TRUE, FALSE, TRUE),     # logical
  zone = factor(c("mixed", "residential", "residential", "industrial", "residential", "recreational"),
                levels = c("residential", "mixed", "industrial", "recreational")),
  dist_km  = c(0.2, 1.4, 1.1, 2.3, 1.8, 2.7)                    # numeric
)

# Inspect
str(samorin_df)
summary(samorin_df)

# Add derived column 'priority' with the same rule
samorin_df$priority <- with(
  samorin_df,
  ifelse(pop > 1000 & !has_playground, "high",
         ifelse(pop > 1000, "medium", "low"))
)
samorin_df$priority <- factor(samorin_df$priority, levels = c("low", "medium", "high"), ordered = TRUE)

# Quick checks
table(samorin_df$zone)
table(samorin_df$priority)

# Show the dataframe
samorin_df
#
```

